Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-17 tex/qccorrespondence.tex
Version: 20210117

\color{blue}

In the case when the system is spatial-homogeneous, the definition of the distribution function is 
\begin{equation}
    f_{n}=\frac{N_{n}}{ \delta \epsilon_{n} },
\end{equation}
and using the same limit we obtain
\begin{eqnarray}
    \Ss&=& \sum_{M=1}^{K} \sum_n
    \left[  
           \frac{N_{n}}{ \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{n}}{ \delta \epsilon_{n}}
           \right)
    \right]  \delta V_M \delta \epsilon_{n} = V \sum_n
    \left[  
           \frac{N_{n}}{ \delta \epsilon_{n}} \ln 
           \left( 
                  \frac{N_{n}}{ \delta \epsilon_{n}}
           \right)
    \right] \delta \epsilon_{n} \nonumber \\
    &=& V \sum_n \left[N_n \ln N_n - N_n \ln \delta \epsilon_n  \right]=\delta V_M K H_{B},\label{h-quantic4}
\end{eqnarray}
where $H_{B}$ now is (\ref{eq:cHdef}) expressed is the form of (\ref{reduce-h}) shown by Tolman.\footnote{See \cite{bib:tolman} eq. (102.6) for more information.}\\
We can observe that if the number of particles is too big compared to the number of cells $K$, then we can conclude that the sum of the local variational entropy is equal to the variational entropy of the entire system.

In the equilibrium case, we have the following variational entropy
\begin{eqnarray}
    \Ss&=& \sum_{M=1}^{K}\sum_{n}
        \left[
                \frac{1}{e^{-\alpha_M-\beta_M \epsilon}\pm 1} \ln 
                    \left(
                            \frac{1}{e^{-\alpha_M-\beta_M \epsilon}\pm 1}
                    \right)
        \right. \nonumber \\
          && \pm \left. \left(
                        1 \mp \frac{1}{e^{-\alpha_M-\beta_M \epsilon}\pm 1}
                  \right) \ln
                \left(
                        1 \mp \frac{1}{e^{-\alpha_M-\beta_M \epsilon}\pm 1}            
                \right) \right]. \label{equilibriumvariational}
\end{eqnarray}
To obtain the classical limit, that is, to recover the Maxwell-Boltzmann distribution from the Fermi-Dirac and Bose-Einstein distribution, it is necessary to hold the following approximation
\begin{equation}
    e^{-\alpha_M-\beta_M \epsilon}\gg 1, \label{classicalapproximation}
\end{equation}
and then
\begin{equation}
    \frac{1}{e^{-\alpha_M-\beta_M \epsilon}\pm 1} \approx e^{\alpha_M+\beta_M \epsilon}.
\end{equation}
If we apply (\ref{classicalapproximation}) to (\ref{equilibriumvariational}), we obtain
\begin{eqnarray}
    \Ss&=& \sum_{M=1}^{K}\sum_{n}
        \left[
                e^{\alpha_M+\beta_M \epsilon} \ln 
                    \left(
                            e^{\alpha_M+\beta_M \epsilon}
                    \right)
        \right. \nonumber \\
          && \pm \left. \left(
                        1 \mp e^{\alpha_M+\beta_M \epsilon}
                  \right) \ln
                \left(
                        1 \mp e^{\alpha_M+\beta_M \epsilon}            
                \right) \right],
\end{eqnarray}
but if (\ref{classicalapproximation}) holds, then 
\begin{equation}
     e^{\alpha_M+\beta_M \epsilon}\ll 1,
\end{equation}
consequently 
\begin{equation}
    \ln(1\mp e^{\alpha_M+\beta_M \epsilon}) \approx \ln 1 \approx 0,
\end{equation}
and finally
\begin{equation}
    \Ss=\sum_{M=1}^{K}\sum_{n}
        \left[
                e^{\alpha_M+\beta_M \epsilon} \ln 
                    \left(
                            e^{\alpha_M+\beta_M \epsilon}
                    \right)
        \right],
\end{equation}
that corresponds to the classical $H$-functional in equilibrium described in the energy space.

Finally, as a complementary topic, in the next section, we propose a method to obtain the quantum analogous from the Boltzmann transport equation for relaxation processes.

>>>>

Finally, the sum over the quatum energy levels can be replaced by an integral over the velocities
by invoking both the uncertainty principle and the fact that, for free particles, the continuum
energy spectrum
can be written as a function of the velocity. Hence the quantum $H$-functional
transforms, in the classical limit, to:
%
\begin{equation}\label{eq:qHclassLim}
    \mathcal{H}= \sum_{M=1}^{K}\int C'f_M(\vec v,t)\ln\big[C' f_M(\vec v,t)\big] d\vec v,
\end{equation}
%
where $C'$ collects the appropriate constants stemming from writing the energy spectrum as
a function of $\vec v$.

Because we have shown that both classical and quantum $H$-functionals, Eqs.~(\ref{eq:cHdef}) and
(\ref{eq:qHdef}) recover the Maxwell-Boltzmann, Bose-Einstein, or Fermi-Dirac distribution
functions in equilibrium, we will omit the quantum classical correspondence of the distribution functions,
as it has been thoroughly discussed in textbooks.

### 2021-01-17 tex/quantum.tex
Version: 20210117

\begin{proof}
To proof the quantum $H$-theorem, we start by taking the time-derivative of Eq. (\ref{eq:qHdef}):
%
\begin{equation}\label{eq:deltaH}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M.
\end{equation}
%
Subsequently, substitute Eq.~(\ref{eq:qFirstOrd}) in the above equation to obtain:
%
\begin{eqnarray}\label{eq:cambioH1}
    \frac{d\mathcal{H} (t)}{dt}&=&
      \sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[
        \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})}
      \right]\dot{g}_{nM} \delta V_M \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{
      \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}
      -\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}
    \right \}\delta V_M.
\end{eqnarray}
%
By representing the logarithmic terms, corresponding to
Fermi and Bose gases, through a Taylor series around $\bar f_n g_{nM}=0$, and
substituting this into Eq.~(\ref{eq:cambioH1}), one obtains:
%
\begin{subequations}\label{eq:qlnApprox}
\begin{equation}\label{eq:qlnApproxFermions}
	\ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]
    	\approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}
\end{equation}
%
and
%
\begin{equation}\label{eq:qlnApproxBosons}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM},
\end{equation}
\end{subequations}
%
respectively. Eq.~(\ref{eq:qlnApproxFermions}) is valid because, for non-extremely degenerated Fermi gases, 
$1-\bar{f}_n \gg \bar{f}_n|g_{nM}|$, and Eq.~(\ref{eq:qlnApproxBosons}) is fulfilled because,
for Boson gases,
$1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$ when $\bar{f}_n \gg \bar{f}_n |g_{nM}|$.

@@

Substitute the previous expression in Eq.~(\ref{eq:cambioH3}) to obtain:
%
\begin{equation}\label{eq:cambioH4}
   \frac{d\mathcal{H}}{dt}=\sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}
   \sum_M  g_{nM}\dot{g}_{nM} \delta V_M \leq 0.
\end{equation}
\end{proof}
Thus, the time evolution of the system occurs in such a way that 
$\frac{d\mathcal{H}}{dt}$ is a decreasing function toward zero.
%

>>>>

%\begin{proof}
To proof the quantum $H$-theorem, we start by taking the time-derivative of Eq. (\ref{eq:qHdef}):
%
\begin{equation}\label{eq:deltaH}
   \frac{d \mathcal{H} (t)}{dt}= \sum_n \sum_{M=1}^{K} \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M.
\end{equation}
%
Subsequently, substitute Eq.~(\ref{eq:qFirstOrd}) in the above equation to obtain:
%
\begin{eqnarray}\label{eq:cambioH1}
    \frac{d\mathcal{H} (t)}{dt}&=&
      \sum_n \sum_{M=1}^{K} \bar{f}_{n}\ln \left[
        \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})}
      \right]\dot{g}_{nM} \delta V_M \nonumber \\
    &=&\sum_n \sum_{M=1}^{K} \bar{f}_n \left \{
      \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}
      -\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}
    \right \}\delta V_M.
\end{eqnarray}
%
By representing the logarithmic terms, corresponding to
Fermi and Bose gases, through a Taylor series around $\bar f_n g_{nM}=0$, and
substituting this into Eq.~(\ref{eq:cambioH1}), one obtains:
%
\begin{subequations}\label{eq:qlnApprox}
\begin{equation}\label{eq:qlnApproxFermions}
	\ln[1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]
    	\approx \ln[1\mp\bar{f}_n]\mp\frac{\bar{f}_n}{1\mp\bar{f}_{n}} g_{nM}
\end{equation}
%
and
%
\begin{equation}\label{eq:qlnApproxBosons}
    \ln [\bar{f}_n+\bar{f}_n g_{nM}] \approx \ln [\bar{f}_n]+ g_{nM},
\end{equation}
\end{subequations}
%
respectively. Eq.~(\ref{eq:qlnApproxFermions}) is valid because, for non-extremely degenerated Fermi gases, 
$1-\bar{f}_n \gg \bar{f}_n|g_{nM}|$, and Eq.~(\ref{eq:qlnApproxBosons}) is fulfilled because,
for Boson gases,
$1+\bar{f}_n \gg \bar{f}_n |g_{nM}|$ when $\bar{f}_n \gg \bar{f}_n |g_{nM}|$.

@@

Substitute the previous expression in Eq.~(\ref{eq:cambioH3}) to obtain:
%
\begin{equation}\label{eq:cambioH4}
   \frac{d\mathcal{H}}{dt}=\sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}
   \sum_M  g_{nM}\dot{g}_{nM} \delta V_M \leq 0.
\end{equation}
%\end{proof}
Thus, the time evolution of the system occurs in such a way that 
$\frac{d\mathcal{H}}{dt}$ is a decreasing function toward zero.
%
